---
title: "Vignette of package bioassays"
author: "Anwar Azad Palakkan, Jamie Davies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{bioassays-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(bioassays)
```

# Introduction

In a cell culture lab various cellular assays are performed. The package
"bioassays" will help to analyse the results of these experiments performed in
multiwell plates. The usage of various functions in the "bioassays" package is
provided in this article.

The functions in this package can be used to summarise data from any multiwell
plate, and by incorporating them in a loop several plates can be analyzed
automatically. Two examples are also provided in the article.

# prerequisite

The output reading from the instrument (eg.spectrophotometer) should be in a
matrix format. An example data file (csv format) is shown below. If the data is
in .xls/.xlsx format `read_excel` function from the `readxl` package can be used.


```{r rawdata}
data(rawdata96)
head(rawdata96)
```

A metadata file is needed for the whole experiment. The `row` and `col` columns
must be present in the metadata file to indicate the location of well. An
example is given below.

```{r metadata}
data(metafile96)
head(metafile96)
```

# Functions

## 1. Function: extract_filename

`extract_filename` helps to extract information from the file name. syntax is
`extract_filename(filename,split = " ",end = ".csv", remove = " ", sep="-")`.
**filename** is the file name. **split** is the portions at which the name has
to be split (default is space " "). **end** is the extension of file name that
need to be removed (default is ".csv"). **remove** is the portion from the file
name that need to be omitted after splitting (default is space " "). **sep** add
a symbol between separate sections, default is "-".

This function is useful for extracting specific information from file names,
like compound name, plate number etc, to provide appropriate analysis.

For e.g.

```{r extract_fname eg1}
extract_filename("L HEPG2 P3 72HRS.csv")
```

```{r extract_fname eg2}
extract_filename("L HEPG2 P3 72HRS.csv", split=" ",end=".csv",remove="L",sep="")
```

## 2. Function: rmodd_summary

`rmodd_summary` helps to remove outliers and summarises the values from a given
set of values. The syntax is `rmodd_summary(x, rm = "FALSE", strict= "FALSE",
cutoff=80,n=3)`. **x** is a numeric vector. **rm = TRUE** if want to remove
outliers. If **strict = FALSE** values above/below 1.5 IQR are omitted (outliers
omitted). If **strict = TRUE** more aggresive outlier removal is used to bring
the %cv below the **cutoff**. **n** is the minimum number of samples you need
per group if more aggressive outlier removal is used.

For example:

```{r rmodd_summary eg1}
x<- c(1.01,0.98,0.6,0.54,0.6,0.6,0.4,3)
rmodd_summary(x, rm = "FALSE", strict= "FALSE", cutoff=80,n=3)
```

```{r ermodd_summary eg2}
rmodd_summary(x, rm = "TRUE", strict= "FALSE", cutoff=80,n=3)
```

```{r rmodd_summary eg3}
rmodd_summary(x, rm = "TRUE", strict= "TRUE", cutoff=20,n=5)
```


## 3. Function: data2plateformat

`data2plateformat` converts the data (eg: readings from a 96 well plate) to an
appropriate matrix format. Syntax is `data2plateformat(data, platetype = 96)`.
**data** is the data to be formatted. **platetype** is the plate from which the
data is coming. It can take 6, 12, 24, 96 or 384 values to represent the
corresponding multiwell plate type.

For e.g. To rename columns and rows of `rawdata96` to right format.


```{r data2plateformat eg2 }
rawdata<-data2plateformat(rawdata96,platetype = 96)
head(rawdata)
```

## 4. Function: plate2df

`plate2df` coereces a matrix with values from multiwell plates into data.frame.
The function uses column names and row names of `datamatrix` (2D data of a
mutliwell plate) and generates a dataframe with row, col (column) and position
indices. The `value` column represents the corresponding value in the
`datamarix`.

Syntax is `plate2df(datamatrix)`. **datamatrix** is the data in matrix format. 

For eg.

```{r plate2df eg1}
OD_df <- plate2df(rawdata)
head(OD_df)
```

## 5. Function: matrix96

`matrix96` helps to convert a data.frame into a matrix. The syntax is
`matrix96(dataframe,column,rm="FALSE")`. **dataframe** is the data.frame to be
formatted. The data.frame must cotain `row` and `col` columns. **column** is the
name of column that needs to be converted into a matrix. If **rm= "TRUE"** then
-ve and NA are set to 0.


For e.g.

```{r matrix96 eg2}
matrix96(OD_df,"value")
```

```{r matrix96 eg3}
matrix96(OD_df,"position")
```

## 6. Function: plate_metadata

`plate_metadata` combines the plate-specific information (like compound used,
standard concentration, dilution of samples, etc) and metadat into  unique plate
metadata. The syntax is `plate_metadata(plate_details,
metadata,mergeby="type")`. **plate details** is the plate specific information
that needs to be added to metadata. **metadata** is the metadata for the whole
experiment. **mergeby** is the column that is common to both metadata and
`plate_meta` (this column will be used to merge the information).

For eg. An incomplete meta data

```{r plate_metadata eg 2 }
head(metafile96)
```

Plate specific details are.

```{r plate_metadata eg 3, echo=TRUE}
plate_details <- list("compound" = "Taxol",
                "concentration" = c(0.00,0.01,0.02,0.05,0.10,1.00,5.00,10.00),
                "type" = c("S1","S2","S3","S4","S5","S6","S7","S8"),
                "dilution" = 1)
```


Using the plate-specific info, the metadata can be filled by calling 
`plate_metadata` function.


```{r plate_metadata eg 5, message=FALSE, warning=FALSE}
plate_meta<-plate_metadata(plate_details,metafile96,mergeby="type")
head(plate_meta)
```

To join both plate_meta and OD_df, the `dplyr::inner_join` function can be used.


```{r plate_metadata eg 6, message=FALSE, warning=FALSE}
data_DF<- dplyr::inner_join(OD_df,plate_meta,by=c("row","col","position"))

head(data_DF)
```


## 7. Function: heatplate

`heatplate` creates a heatmap representation of a multiwell plate. The syntax is `heatplate(datamatrix,name,size=7.5)`. **datamatrix** is the data in matrix
format. An easy way to create this is by calling the `matrix96` function, as 
explained before. **name** is the name to be given for heatmap, 
**size** is the size of each well in the heatmap (default is 7.5). 

This function will produce a heatmap of normalized values if the `variable` is 
numeric. If it is a factor, it will simple provide a coloured  categorical plot.

eg 1. Categorical plot

```{r heatplate eg:1, warning= FALSE}
datamatrix<-matrix96(metafile96,"id")
datamatrix
```


```{r heatplate eg:2, warning= FALSE, fig.width=4, fig.height=3.5}
heatplate(datamatrix,"Plate 1", size=5)
```

eg 2. Heatmap

```{r heatplate eg:3, warning=FALSE}
rawdata<-data2plateformat(rawdata96,platetype = 96)
OD_df<- plate2df(rawdata)
data<-matrix96(OD_df,"value")
data
```


```{r heatplate eg:4, warning=FALSE, fig.width=4, fig.height=3.5}
heatplate(data,"Plate 1", size=5)
```


## 8. Function: reduceblank

`reduceblank` helps to reduce blank values from the readings.

The syntax is `reduceblank (dataframe,x_vector,blank_vector,y)`. **dataframe**
is the data. **x_vector** are the entries for which the blank has to be
subtracted If all entries have to subtracted, then use "All". x_vector should be
a vector eg: c("drug1","drug2",drug3" etc). **blank_vector** is a vector of
blank names whose value has to be subtracted eg:
c("blank1","blank2","blank3","blank4")). This function will subtract the first
blank vector element from first x_vector element and so on. **y** is the column
name where the action will take place. y should be numeric. The results will
appear as a new column named `blankminus`.

For eg.

```{r reduceblank eg1, warning=FALSE}
data_DF<-reduceblank(data_DF, x_vector =c("All"),
                     blank_vector = c("Blank"), "value")
head(data_DF)

```

## 9. Function: estimate

`estimate` estimates an unknown variable (eg: concentration) based on the
standard curve. Syntax is
`estimate(data=dataframe,colname="blankminus",fitformula=fit,
methord="linear/nplr")`. **data** is the dataframe which needs to be evaluated.
**colname** is the column name for which the values has to be estimated.
**fitformula** represents the model to be fit **methord** specifies if linear or
nonparametric logistic curve was used for the fitformula.

For eg: data_DF is a dataframe for which the concentration has to be estimated based on the value of blankminus.

For filtering the `standards`

```{r standards}
std<- dplyr::filter(data_DF, data_DF$id=="STD")  
std<- aggregate(std$blankminus ~ std$concentration, FUN = mean )
colnames (std) <-c("con", "OD")
head(std)
```

To fit a standard curve:

fit1 is the 3 parameter logistic curve model and fit2 is the linear regression
model. The appropriate one for your experiment can be used.

```{r fitmodels}
fit2<-stats::lm(formula = con ~ OD,data = std)# linear model
fit1<-nplr::nplr(std$con,std$OD,npars=3,useLog = FALSE)#  nplr, 3 parameter model
```

For estimating the concentration using linear model

```{r nplr estimating nplr, message=FALSE, warning=FALSE}
estimated<-estimate(data_DF,colname="blankminus",fitformula=fit2,method="linear")
head(estimated)
```

For estimating the concentration using nplr methord

```{r nplr estimating linear, message=FALSE, warning=FALSE}
estimated2<-estimate(data_DF,colname="blankminus",fitformula=fit1,method="nplr")
head(estimated2)
```

## 10. Function: dfsummary

`dfsummary()` summarizes the data.frame (based on a column). It has additional
controls togroup samples and to omit variables not needed. The syntax is
`dfsummary(dataframe,y,grp_vector,rm_vector,nickname,rm="FALSE",param)`.
**dataframe** is the data. **y** is the numeric variable (column name) that has
to be summarized. **grp_vector** is a vector of column names, determining which
samples will be grouped. The order of the elements in grp_vector determines the
order of grouping. **rm_vector** is the vector of items to be omit before
summarizing. **nickname** is the name for the output data.frame. Set
**rm="FALSE"** if outliers should not be removed. To apply more stringent
thresholds for removing outliers, parameters can be provided in the  **param**
vector. **param** has to be entered in the format
`c(strict="TRUE",cutoff=40,n=12)`. For details please refer to the
`rmodd_summary `function.

In the following example, the data has to be summarized based on the "type"
column. "estimated" values are summarized. Samples are grouped as per "id".
"STD" and "Blank" values need to be omitted. Outliers are retained (rm="FALSE").
The nickname for the plate is "plate1".

```{r summary 3, message=FALSE, warning=FALSE, out.width=40}
result<-dfsummary(estimated,"estimated",c("id","type"),
        c("STD","Blank"),"plate1", rm="FALSE",
        param=c(strict="FALSE",cutoff=40,n=12))

result
```

## 11. Function: pvalue

The `pvalue()` function applies a t-test to the result dataframe. The syntax is
`pvalue(dataframe,control,sigval)`. **dataframe** is the result of dfsummary.
**control** is the group that is considered as control, **sigval** is the pvalue
cutoff (a value below this is considered as significant).

For example:

```{r pvalue eg1, warning=FALSE, message = FALSE}
pval<-pvalue(result, control="S8", sigval=0.05)
head(pval)
```

